Short-term synaptic and neuronal plasticity act in synergy for deviance detection in spiking networks

### 2022-12-15 paper-1/manuscript.tex
Rev Results:STD (F8)

\textbf{Fig. 6} Contrast between the ablated model with only TA (shown in Figure \ref{fig5}) and the full model with both STD and TA. \textbf{a}, \textbf{b} Data shown in Figure \ref{fig5}ab, dev and msc columns, subtracted from the corresponding data for the full model. \textbf{c} Equivalent contrast (full - ablated model) in the response to A, in columns as above. \textbf{d} Network-wide mean $V_{TA}$ before A trials across networks. Left: Ablated and full model averages. Right: Difference between these model conditions. With the introduction of STD, $V_{TA}$ decreased more in dev than in msc (t = -3.49, p = 0.000342). \textbf{e} Mean number of spikes per target trial across networks. Left: Ablated and full model averages. Right: Difference between these model conditions. With the introduction of STD, the response decreased more in the std sequence than msc (t = -1.9, p = 0.0297), and more in msc than in dev (t = -2.99, p = 0.00168)

However, we noticed a systematic difference in the reduction of both response magnitude and $V_{TA}$ between the dev and msc conditions: The non-target response in the dev sequence was reduced more than that in the msc sequence, leading to a larger difference between the responses in these two contexts. $V_{TA}$ before A trials followed suit, leading to stronger de-adaptation and, as a consequence, a larger A trial response in dev in large parts of the network in the full model compared to TA-only. Across networks, only few cases of an increased dev response were seen; however, the response to the target stimulus (Figure \ref{fig6}e) was reduced least in dev, and most in std. This differential reduction was directly responsible for the larger DDI reported in Figure \ref{fig3}, but how did it arise?

Due to its relatively short time scale, the direct effect of STD across trials was likely limited to repeated trials of the same stimulus, such as the frequent stimulus in oddball sequences. However, by reducing the standard (i.e., the dev non-target) response in the full model, we hypothesized that STD also affects adaptation levels, causing de-adaptation and thereby indirectly increasing the target dev response.

In direct support of the notion that STD indeed carried over between (repeated) trials, we noted first that the response size in the STD-only condition was slightly less in std than in msc or dev (cf. Figure \ref{fig3}c), indicating some amount of SSA. To show that the across-trial effect of STD in the full model was limited to the dev non-target trials, however, we needed a more sophisticated tool, showing the level of depression experienced by each neuron. Since the effect of STD is to reduce excitatory postsynaptic potentials (EPSP), we estimated this reduction per synapse in a given trial type $S$ (e.g. A dev) as the product of the expected number of transmitted spikes $R^S$, the peak EPSP of the synapse ($k = 1.4 mV$, recall that all weights are 1), and its average depression coefficient $(1-x)$ at the start of the trial. Summing these estimates onto each postsynaptic neuron, we arrived at a rough estimate

\begin{equation}
    D^S = \sum_{i \in pre} R^S_i k (1-x_i^{t=0})
\end{equation}

of the average depolarization lost due to STD.

\begin{figure*}%
    \centering
    \includegraphics{fig7.pdf}
    \caption{\textbf{a} Estimated average STD-derived postsynaptic depression $D$ in target (A) trials (top row) and B trials (bottom row). Note that the difference between A and B in msc (middle column) is due only to the network configuration. \textbf{b} Statistics across networks of the contrast (dev - msc) in $D$ (cf. \textbf{a}, right column), and in $V_{TA}$ at the start of A trials (cf. Figure \ref{fig6}d). In each network, averages were taken over neurons with at least 0.2 spikes per dev trial (B trials for $\Delta D^B$, A trials for $\Delta D^A$ and $\Delta V_{TA}$). Whereas the third box ($\Delta V_{TA}$ total) indicates the averages in the full model, the final box ($\Delta V_{TA}$ added) indicates the difference between the full and the ablated (TA-only) model. This additional amount of $\Delta V_{TA}$, caused indirectly by the addition of STD, was less (i.e., more negative) than the directly STD-related $\Delta D^A$ itself (t = -2.09, p = 0.0195)}
    \label{fig7}
\end{figure*}

\textbf{Fig. 7 a} Estimated average STD-derived postsynaptic depression $D$ in target (A) trials (top row) and B trials (bottom row). Note that the difference between A and B in msc (middle column) is due only to the network configuration. \textbf{b} Statistics across networks of the contrast (dev - msc) in $D$ (cf. \textbf{a}, right column), and in $V_{TA}$ at the start of A trials (cf. Figure \ref{fig6}d). In each network, averages were taken over neurons with at least 0.2 spikes per dev trial (B trials for $\Delta D^B$, A trials for $\Delta D^A$ and $\Delta V_{TA}$). Whereas the third box ($\Delta V_{TA}$ total) indicates the averages in the full model, the final box ($\Delta V_{TA}$ added) indicates the difference between the full and the ablated (TA-only) model. This additional amount of $\Delta V_{TA}$, caused indirectly by the addition of STD, was less (i.e., more negative) than the directly STD-related $\Delta D^A$ itself (t = -2.09, p = 0.0195)

In Figure \ref{fig7}, we show that, in A dev and A msc trials, depression had relatively evenly distributed effects across the network. Neither sequence seemed to have a stronger overall effect, despite a slight trend towards weaker depression near the stimulus site in the dev sequence. In sharp contrast, in B dev trials, depression was very strong and very clearly localized to the stimulation site. This location dependence was closely mirrored in the B dev network response, which was largely confined to the area near the stimulation site (data not shown). To quantify this result while avoiding spurious events, we averaged the depression value $D$ over neurons with at least 1 spike per 5 trials of the corresponding type in the dev sequence. Across networks, we found that depression in the dev sequence was systematically stronger for B than for A (t = 7.5, p = 7.07e-12). More importantly, the level of depression in A trials changed considerably less between msc and dev sequences when compared to B trials (t = -6.7, p = 3.92e-10). Taken together, this shows that, indeed, the dominant direct effect of STD was to weaken the response to the frequently presented standard stimulus.

Finally, to complete the picture and confirm our hypothesis in full, we compared the small, but significant relief from depression in A dev trials relative to msc (t = -3.2, p = 0.001) to the difference in $V_{TA}$, applying the same neuron exclusion criteria. We found that the relief from suppression, which was shown in the TA-only analysis above to be a key ingredient to deviance detection, was driven almost exclusively by lower $V_{TA}$ (t = 11, p = 4.06e-19). More importantly, adding STD to the model induced a much larger additional difference in $V_{TA}$ compared to the TA-only condition, than in directly STD-related $D$ (t = -2.09, p = 0.0195), indicating that, indeed, STD affected the deviant response primarily through indirect effects.

>>>>

This naturally raises the question of why, with the addition of STD, the non-target dev response was reduced more than other responses. We hypothesized that, due to its relatively short time constant, the direct effect of STD across trials was limited to repeated trials of the same stimulus, such as the frequent stimulus in oddball sequences.
To show this in the sample network, we correlated the number of spikes in dev target trials with the spike-triggered depression variable $x_j$ at the start of later trials (cf. Equation \ref{eqn-xsyn}) across excitatory neurons. We found that, while the correlation was very clear in the trial immediately following target (Pearson's \rho = -0.984, p = 0), it was almost completely lost by the next trial (\rho = -0.0525, p = 0.138) under the influence of both decay and unrelated activity in the intervening trial. In fact, even without intervening activity, inputs to $x_j$ would decay to a negligible $e^\frac{1 s}{\tau_x} = 0.0013$ within two trials.
Thus, STD could not have any stimulus-specific effects across dev target trials or msc trials.

However, given the substantial correlation from one trial to the next, the frequently presented stimulus B could have experienced stimulus-specific adaptation as a result of STD. Referring back to Figure \ref{fig3}b, we see that there was indeed a significant, if small, effect of presentation frequency even in the ablated model with STD only, yielding a positive SSA index on average.

To show that STD carry-over was responsible for the reduction of non-target dev pulses relative to msc, we needed to estimate the extent to which activity of a given neuron (or, alternatively, its membrane potential $V_m$) is affected by changes in STD. Unlike $V_{TA}$, which directly affects the neuron to which the value belongs, $x_j$ exerts its effects at the other side of a synapse. Therefore, we devised a measure $D$ to estimate the postsynaptic impact of STD in terms of membrane potential lost. $D$ is necessarily activity-dependent -- a postsynaptic neuron's membrane potential is driven only by presynaptic partners that fire -- so must be calculated based on observed activity. Here, we used the average trial response $R_j^S$ of neuron $j$ to a given stimulus $S$, in number of spikes, as the presynaptic drive. Before depression, a single excitatory presynaptic spike evokes a postsynaptic potential (EPSP) peaking at approximately $k = 1.4$ mV at rest (recall that $w = 1$ for all synapses), and STD linearly scales this EPSP by presynaptic $x_j$. Thus, ignoring added depression due to repeated spikes within a trial, and instead using the average depression at the start of trials $x_j^S$, we estimated the average peak EPSP evoked by a presynaptic neuron $j$ as $R_j^S k x_j^S$. Finally, with $x_j \in [0,1]$ and summing over presynaptic neurons, we could estimate the peak EPSP lost to depression in a given neuron $i$ as
\begin{equation}
    D_i^S = \sum_{j \in pre} R_j^S k (1-x_j^S)
\end{equation}
Notice that $D$ is measured in volts due to scaling with $k$, and is therefore roughly comparable to $V_{TA}$, albeit more approximate.

\begin{figure*}%
    \centering
    \includegraphics{fig8.pdf}
    \caption{}
    \label{fig8}
\end{figure*}
\textbf{Fig. 8} Estimated STD-mediated depression $D$ has small direct effects.
\textbf{a} Average depression $D$ in target (A) trials (top row) and B trials (bottom row) in the dev (left) and msc sequence (center), and their contrast (right). The color scale is shared across columns (and between dev and msc) for direct comparability.
\textbf{b} Statistics across networks and stimuli of the contrast (dev - msc) in $D$ ($D^B$ is shown first, since all other data in this panel relate directly to A), and in $V_{TA}$ at the start of A trials (cf. Figure \ref{fig7}d). In each network, averages were taken over neurons active (>0.2 spikes/trial) in either dev or msc trials (B trials for $\Delta D^B$, A trials for $\Delta D^A$ and $\Delta V_{TA}$). Whereas the third box ($\Delta V_{TA}$ total) indicates averages in the full model, the final box ($\Delta V_{TA}$ added) indicates the difference between the full and the ablated (TA only) model

As noted above, we hypothesized that, in the full model, STD is directly responsible only for a reduction of non-target dev trials, relative to msc, whereas differences between target dev and msc trials are not a result of STD directly. To show this, we calculated $D$ for A and B trials, and contrasted it across dev and msc conditions as with $V_{TA}$ before. 
In Figure \ref{fig8}a, we see that target trials (top) experienced depression across large parts of the network in both conditions. While there were differences between the two conditions, they were relatively small and did not appear to follow a readily identifiable pattern. Conversely, B trials (bottom) were spatially more clustered around the stimulation site in both conditions. In the msc sequence, we can attribute the difference between A and B primarily to the response size: The context was the same, but the B response tended not to spread out as much in this network. However, in the dev sequence, the B response was limited almost exclusively to a very local patch around the stimulation site (see also Figure \ref{fig6}a), with depression, of course, similarly limited. As a result, the contrast (dev - msc) shows slightly reduced $D$ in the periphery, but strikingly elevated $D$ near the stimulation site, indicative of increased synaptic depression in dev.

In order to show that this was a robust effect, we turn to statistics across networks and stimuli. We calculated the trial-average depression values $D^A$ and $D^B$ in dev and msc conditions, subtracted them, then calculated the mean over neurons, excluding neurons with less than 0.2 spikes per trial in both conditions [because $\Delta D^B$ $\approx$ zero if I don't do this. Not sure how to justify...]. The resulting contrasts are summarized in the left half of Figure \ref{fig8}b. We note first that, using the same exclusion criteria, we found that neurons received more depression in B trials than in A trials in the dev sequence (t = -4.4, p = 1.17e-05; one-sided t-test, n = 120), but not in the msc sequence (t = -1.14, p = 0.129). Furthermore, $D^B$ was greater in dev than msc (t = 2.94, p = 0.002). Both findings confirm that STD carried over between the frequent standard trials, but not, or less so, between control trials.

However, we also found that $D^A$ was greater in msc than dev (t = -3.4, p = 0.000458), suggesting that STD did exert stimulus-independent suppression in the msc sequence, similar to what we described earlier for TA. The magnitude of the effect, however, was very different. In the right half of Figure \ref{fig8}b, we show the contrast of $V_{TA}$ before A trials, which was clearly much greater than the contrast in $D^A$. We wondered, however, how much $V_{TA}$ was affected by the addition of STD. To assess this, we subtracted the contrast (dev - msc) of $V_{TA}$ in the full model from the same contrast calculated in the ablated model with TA only. This additional decrease in $V_{TA}$ from msc to dev was substantially greater than the corresponding decrease in $D^A$ (t = -1.93, p = 0.0277), indicating that the direct effect of STD on the target dev response was less than indirect effects.

To sum up, adding STD to the model had a number of direct and indirect effects. Firstly, it reduced responses in all conditions, likely by reducing the synaptic output of neurons at stimulus sites, which reliably fired several spikes immediately following stimulation. Secondly, it directly reduced the response to frequently presented dev B trials through stimulus-specific adaptation. Thirdly, STD was very slightly reduced in A trials in dev over msc, mirroring the direct stimulus-independent adaptation effect detailed in the previous section for TA. And finally, the decreased dev B trial response allowed TA in the wider network to recover more, which in turn led to increased responses in dev A trials. This indirect effect was substantially greater than the direct effect on target trials, as shown above, indicating that it was solely or largely responsible for the increased DDI.
